PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2023-02-17 manuscript_00.Rmd
plotting and text updates

The retina is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011]. These six populations can be further divided into dozens of cell types [@yanCellAtlasHuman2020a]. Furthermore, human disease of the eye can derive from damage to specific cell types. Glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the transcriptomics of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

The modern era of single cell transcriptomics has been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

These publications's foci includes, but are not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms. Independent querying of these datasets requires a laborious series of steps.

@@

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries impossible. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. These cutoffs were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is #a99dced There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app. The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp. Data is available for download at plae.nei.nih.gov (click "Data") and the Seurat object and metadata for the resource has been deposited at Zenodo with accession 10.5281/zenodo.7071682.

@@

```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap, echo = FALSE}

@@

study_count_by_tissue <- meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  #filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(Organ = factor(Organ, levels = c('Eye','Body'))) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) +
  cowplot::theme_cowplot() +
  coord_flip() +
  theme(legend.position = "none") +
  scale_y_continuous(expand = c(0, 0))  + xlab('') +
  ggforce::facet_col(~Organ, scales = 'free_y', space = 'free')

@@

cell_counts <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(CellType_predict, organism) %>%
  summarise(Count = n()) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=sCount, label = Count, fill = organism, group = organism)) +
  ylab('log2(Number of Cells)') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

study_counts_by_cell <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(organism, CellType_predict, study_accession) %>%
  summarise(Count = n()) %>% 
  filter(Count > 10) %>% 
  summarise(Count = n()) %>% 
  mutate(Count = case_when(CellType_predict != 'Unlabelled' ~ Count)) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=Count, label = Count, fill = organism, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

@@

colnames(timings) <- c('PLAE',"Spectacle",'Tabula Sapiens\n(CELLxGENE)')

@@

timing_plot <- timings %>% as_tibble(rownames = 'Test') %>% 
  pivot_longer(-Test) %>% 
  mutate(Test = factor(Test, levels = row.names(timings))) %>% 
  mutate(Text = case_when(is.na(value) ~ 'Not\npossible',
                          TRUE ~ '')) %>% 
  ggplot(aes(x=name,y=value, label = Text)) + 
  geom_bar(stat= 'identity') + 
  geom_text(aes(y=0), hjust = 0, angle = 90) +
  facet_wrap(~Test, scales = 'free_y', ncol = 2) + 
  xlab('Resource') + 
  ylab('Seconds to Load') + 
  cowplot::theme_cowplot() +
  scale_y_continuous(breaks = breaks_fun)

ab <- cowplot::plot_grid(cell_counts + theme(legend.position = "none", axis.ticks.y = element_blank()), 
                         study_counts_by_cell + theme(axis.ticks.y = element_blank()), 
                         nrow = 1, 
                         rel_widths = c(1,1), labels = c('a','b'))

cd <- cowplot::plot_grid(study_count_by_tissue + theme(axis.ticks.y = element_blank()), 
                         timing_plot, 
                         nrow = 1, 
                         rel_widths = c(0.7,1),
                         labels = c('c','d'))

cowplot::plot_grid(ab, cd, ncol = 1, rel_heights = c(1,0.8))

@@

The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, integrating the independent studies to deliver cross study comparisons, and, crucially, providing full availability of the underlying data for outside usage.

@@

```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app. A. UMAP view showing *Crx* expression with cell type labels, B. Information table showing expression of Crx across celltypes and organism. C. Box plot of selected genes in a user selected cell types across different studies. D. In silico in-situ of Rho. E. Dotplot demonstating efficient display of gene expression across many cell types.')
knitr::include_graphics('fig/fig1.png')
```

@@

```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms. Rows are split by what cell type the gene is a marker for. Columns are split by the cell types. More intense red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative. Relative number of citations are shown for the gene or gene " and retina" when searching against PubMed compared to 1000 randomly chosen genes.')

@@

```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
neuro_cap <- fig_cap(name='neuro', caption = 'A. Genes proposed by Lu et al. as being significantly differentially expressed during the RPC to neurogenic differentiation in human. We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells. The standard error (calculated by DESeq2) is plotted. Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively. B. Genes from Lu et al. that we replicated with the scEiaD data. C. Novel genes we propose as being differentially expressed between RPC and neurogenic cells')
# tic()
# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
#   tbl("diff_testing") %>% 
#   filter(Against  == "RPC",   
#          Group == 'CellType (Predict)', 
#          Base == "Neurogenic Cell") %>% 
#   group_by(Base, Against, Gene) %>% 
#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
#   arrange(-`mean log2FC`) %>%   
#   collect() %>% 
#   filter(!is.na(`mean log2FC`))
# 
# 
# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
# # left_join(gene_tab, by = 'GENEID') %>% 
# # relocate(SYMBOL, GENEID) %>% 
# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
# # arrange(-`mean log2FC`)
#

@@

exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)

input$exp_plot_genes <- up[up %in% g_lu]
exp_plot_existing <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
a_legend <- get_legend(exp_plot_new)
neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + xlab(''),
                                  exp_plot_new + theme(legend.position="none") + ylab('') + xlab(''), 
                                  labels = c('b','c'),
                                  rel_widths = c(0.6,1)),
                        a_legend, ncol = 1, rel_heights = c(1,0.2))

@@

```{r undersupported_genes, fig.height=23, fig.width=8, echo = FALSE, fig.cap = undersupported_genes_cap}
undersupported_genes_cap <- supFig_cap(name = 'undersupported_genes', caption = 'Well supported genes with a mean log2FC > 5 and zero pubmed citation hits for [gene] AND Retina search.')

>>>>

The outer layer of the eye consists of the cornea and sclera. Beneath the sclera is the uvea which is made of the iris, ciliary body, and the choroid. The inner most layer of the eye is the retina, which is backed by the retinal pigmented epithelium (RPE). The retina alone is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011]. These six populations can be further divided into dozens of cell types [@yanCellAtlasHuman2020a]. Furthermore, human disease of the eye can derive from damage to specific cell types. Glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the transcriptomics of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

The modern era of single cell transcriptomics starts with the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment [@kleinDropletBarcodingSingleCell2015]. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since the Macosko publication. An incompmlete list of what the ocular community has used single cell technology across dozens of studies would include studing developmental transcriptomic dynamics, identifying genes to distinguish cell types, proposing finer gradations of cell types within existing categories, comparing and contrasting ocular transcriptomes between organisms, and identify differences in spatially different cells.

@@

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast,the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries impossible. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. These cutoffs were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is #a99dced. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app. The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp. Data is available for download at plae.nei.nih.gov (click "Data") and the Seurat object and metadata for the resource has been deposited at Zenodo with accession 10.5281/zenodo.7071682.

@@

```{r, fig.height=13, fig.width=11, fig.cap = scEiaD_data_overview_cap, echo = FALSE}

@@

study_count_by_tissue <- meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  #filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(Organ = factor(Organ, levels = c('Eye','Body'))) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) +
  cowplot::theme_cowplot() +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0))  + xlab('') +
  ggforce::facet_col(~Organ, scales = 'free_y', space = 'free')

@@

cell_counts <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(CellType_predict, organism) %>%
  summarise(Count = n()) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=sCount, label = Count, fill = organism, group = organism)) +
  ylab('log2(Number of Cells)') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

study_counts_by_cell <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(organism, CellType_predict, study_accession) %>%
  summarise(Count = n()) %>% 
  filter(Count > 10) %>% 
  summarise(Count = n()) %>% 
  mutate(Count = case_when(CellType_predict != 'Unlabelled' ~ Count)) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=Count, label = Count, fill = organism, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

@@

colnames(timings) <- c('PLAE',"Spectacle",'Tabula\nSapiens\n(CELLxGENE)')

@@

timing_plot <- timings %>% as_tibble(rownames = 'Test') %>% 
  pivot_longer(-Test) %>% 
  mutate(Test = factor(Test, levels = row.names(timings))) %>% 
  mutate(Text = case_when(is.na(value) ~ 'Cannot',
                          TRUE ~ '')) %>% 
  ggplot(aes(x=name,y=value, label = Text)) + 
  geom_bar(stat= 'identity') + 
  geom_text(aes(y=0), hjust = 0, angle = 90) +
  facet_wrap(~Test, scales = 'free_y', ncol = 2) + 
  xlab('Resource') + 
  ylab('Seconds to Load') + 
  cowplot::theme_cowplot() +
  scale_y_continuous(breaks = breaks_fun)

ab <- cowplot::plot_grid(cell_counts + theme(legend.position = "none", axis.ticks.y = element_blank()), 
                         study_counts_by_cell + theme(legend.position = "none", axis.ticks.y = element_blank()), 
                         nrow = 1, 
                         rel_widths = c(1,1), labels = c('a','b'))

cd <- cowplot::plot_grid(study_count_by_tissue + theme(legend.position = c(0.4, 0.3)) + theme(axis.ticks.y = element_blank()), 
                         timing_plot, 
                         nrow = 1, 
                         rel_widths = c(0.7,1),
                         labels = c('c','d'))

cowplot::plot_grid(ab, cd, ncol = 1, rel_heights = c(1,0.85))

@@

The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing the faw sequence data in a consistent framework, batch correcting the independent studies, and, crucially, providing full availability of the underlying data for outside usage.

@@

```{r PLAE_screenshots, fig.height = 10, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app. A. UMAP view showing *Crx* expression with cell type labels, B. Information table showing expression of Crx across celltypes and organism. C. Box plot of selected genes in a user selected cell types across different studies. D. In silico in-situ of Rho. E. Dotplot demonstating efficient display of gene expression across many cell types.')
knitr::include_graphics('fig/figure2 copy.png')
```

@@

```{r consist_genes_heatmap, fig.height=23, fig.width=12, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms. Rows are split by what cell type the gene is a marker for. Columns are split by the cell types. More intense red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative. Relative number of citations are shown for the gene or gene " and retina" when searching against PubMed compared to 1000 randomly chosen genes.')

@@

```{r neuro, fig.width=16, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
neuro_cap <- fig_cap(name='neuro', caption = 'A. Genes proposed by Lu et al. as being significantly differentially expressed during the RPC to neurogenic differentiation in human. We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells. The standard error (calculated by DESeq2) is plotted. Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively. B. Genes from Lu et al. that we replicated with the scEiaD data. C. Novel genes we propose as being differentially expressed between RPC and neurogenic cells')
# tic()
# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
#   tbl("diff_testing") %>% 
#   filter(Against  == "RPC",   
#          Group == 'CellType (Predict)', 
#          Base == "Neurogenic Cell") %>% 
#   group_by(Base, Against, Gene) %>% 
#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
#   arrange(-`mean log2FC`) %>%   
#   collect() %>% 
#   filter(!is.na(`mean log2FC`))
# 
# 
# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
# # left_join(gene_tab, by = 'GENEID') %>% 
# # relocate(SYMBOL, GENEID) %>% 
# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
# # arrange(-`mean log2FC`)
#

@@

exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + theme(legend.box="vertical", legend.justification="center")

input$exp_plot_genes <- up[up %in% g_lu]
exp_plot_existing <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
a_legend <- get_legend(exp_plot_new)
neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + ylab("CellType (Predict)") ,
                                  exp_plot_new + theme(legend.position="none", axis.text.y =element_blank(), axis.ticks.y = element_blank()) + ylab('') , 
                                  labels = c('b','c'),
                                  rel_widths = c(0.6,1)),
                        a_legend, ncol = 1, rel_heights = c(1,0.2))

@@

```{r undersupported_genes, fig.height=23, fig.width=12, echo = FALSE, fig.cap = undersupported_genes_cap}
undersupported_genes_cap <- supFig_cap(name = 'undersupported_genes', caption = 'Well supported genes with a mean log2FC > 5 and zero pubmed citation hits for [gene] AND Retina search.')
